High Resolution Identification of Protein-DNA Binding Events and Quality Control for ChIP-exo data

### 2015-11-12 rmarkdown/chip_exo_paper.Rnw
finished complexity analysis, and added comments to paper

DNA libraries generated by the ChIP-exo protocol seem to be less
complex than the libraries generated by ChIP-Seq
\cite{exo_review}. Hence, most of current guidelines \cite{encode_qc}
may not be aplicable on ChIP-exo. To address this challenge, we
suggest a series of quality control visualizations to understand which
biases are present in ChIP-exo data. Related to quality control,
Previous ChIP-exo analysis used ChIP-Seq samples to compare the
resolution between experiments (\cite{exo1}, \cite{exo2},
\cite{exoillumina}). In \cite{carroll.qc}, Carroll et. al. studied the
use of the Strand-Cross Correlation (SCC) \cite{strandcc}. and showed
that by filtering blacklisted regions the estimation of the SCC is
improved. However, this method requires to know blacklisted regions in
advance which may not be available. In our pipeline we propose two
out-the-shelf metrics equivalent to RSC and NSC for the estimation of
the signal-to-noise ratio for a ChIP-exo sample.

@@

\begin{table}[h]
  \centering
\begin{tabular}{l|r|r|r|r|r|r|r}
\hline\hline
Condition & Rep. & Nr. reads & PBC & SSD & FSR & NSC & RSC \\
\hline\hline
Aerobic & 1 & 1,454,566 & 0.2156 & 0.0830 & 0.5012 & 114.14 & -1.670\\
\hline
Aerobic & 2 & 864,714 & 0.2471 & 0.0488 & 0.4923 & 160.93 & -1.503 \\
\hline
Rif-0min & 1 & 960,256 & 0.2823 & 0.0361 & 0.4998 & 10.29 & -0.092\\
\hline
Rif-0min & 2 & 2,247,295 & 0.2656 & 0.1091 & 0.5047 &  25.08 & -0.2279 \\
\hline
Rif-20min & 1 & 1,940,387 & 0.2698 & 0.0820 & 0.5054 & 17.69 & -0.450 \\
\hline
Rif-20min & 2 & 4,229,574 & 0.2153 & 0.1647 & 0.5071 &  14.11 & -0.533\\
\hline
\end{tabular}  
\caption{Usual quality control indicators applied to our $\sigma^{70}$ samples. PBC stands PCR-bottleneck coefficient, SSD for standarized standard deviation, FSR for forward strand ratio, NSC for normalized strand cross-correlation coefficient and RSC for relative strand cross-correlation coefficient.}
  \label{tab:qcbase}
\end{table}

@@

Figure \ref{fig:qcdiagram} shows a flowchart for the ChIPexoQC
pipeline. Which basically partitions genome by keeping the
non-diggested ChIP-exo regions. Then, for each region it calculates a
series of independent statistics. Finally, it creates several
visualizations that were designed to asses the overall quality levels
of A ChIP-exo sample.

@@

\subsubsection{Assessing the strand imbalance in ChIP-exo data}
\label{sec:strand_imbalance}

The strand imbalance assesment is based in the observation that the
enriched regions usually have a higher concentration of fragments,
therefore we examine the FSR (defined as the ratio of the number of
forward stranded reads divided by the total number of reads in a given
region) as the region with lower depth are being filtered out. This
indicator is of particular importance, since several methods rely on
the ``peak-pair'' assumption. In table \ref{tab:qcbase}, we calculated
the FSR and noticed that for all the samples, it's value is close to
0.5, which means that there are roughly the same amount of reads in
both strands. However, as we can see in figure \ref{fig:fsr}

@@

We first compared various factors that could affect binding site
identification between ChIP-exo and ChIP-Seq data. in order to compare
distribution of signal and background between ChIP-exo and ChIP-Seq
data, we calculated ChIP tag counts accross the genome by counting the
number of reads mapping to each of \Sexpr{bin_size} non-overlapping
window after extending reads by \Sexpr{frag_len} to their 3' end
directions. ChIP tag counts in ChIP-exo data were linearly related to
ChIP tag counts in ChIP-Seq data for the regions with high ChIP tag
counts (Figure \ref{fig:comp}A). This implies that signals for
pontential binding sites are well reproducible between ChIP-exo and
ChIP-Seq data. On the other hand, there was clear difference in the
background distribution between them. In CHIP-Seq datam reads were
almost uniformly distributed over background (non-binding) regions and
the ChIP tag counts in there regions were significantly larger than
zero. In contrast, in ChIP-exo data, there was larger variation in
ChIP tag counts among background regions and ChIP tag counts were much
lower in these regions compared to ChIP-Seq data. There were also
large proportion of regions without any read in ChIP-exo data. There
results indicate that background distribution of CHIP-exo data is less
homogeneous than that of ChIP-Seq data.

\begin{figure}[h]
  \centering
  \includegraphics[width = .6\textwidth,page = 3 ]{../figs/for_paper/ChIPseqPET_ChIPexo_tagCount_comparison.pdf}
  \includegraphics[width = .9\textwidth ]{../figs/for_paper/MA_plot_fwd_bwd_by_seq.pdf}
  \caption{Comparison}
  \label{fig:comp}
\end{figure}

@@

@@

@@

>>>>

DNA libraries generated by the ChIP-exo protocol seem to be less
complex than the libraries generated by ChIP-Seq
\cite{exo_review}. Hence, most of current guidelines \cite{encode_qc}
may not be aplicable on ChIP-exo, additionally to our knowledge there
are not established quality control pipelines for ChIP-exo. To address
this challenge, we suggest a series of quality control visualizations
to understand which biases are present in ChIP-exo data. Related to
quality control, Previous ChIP-exo analysis used ChIP-Seq samples to
compare the resolution between experiments (\cite{exo1}, \cite{exo2},
\cite{exoillumina}). In \cite{carroll.qc}, Carroll et. al. studied the
use of the Strand-Cross Correlation (SCC) \cite{strandcc}. and showed
that by filtering blacklisted regions the estimation of the SCC is
improved. However, this method requires to know blacklisted regions in
advance which may not be available. In our pipeline we propose two
out-the-shelf metrics equivalent to RSC and NSC for the estimation of
the signal-to-noise ratio for a ChIP-exo sample.

@@

\begin{table}[h]
  \centering
\begin{tabular}{l|r|r|r|r|r|r}
\hline\hline
Condition & Rep. & Nr. reads & PBC & SSD & FSR & NSC  \\
\hline\hline
Aerobic & 1 & 1,454,566 & 0.2156 & 0.0830 & 0.5012 & 114.14 \\
\hline
Aerobic & 2 & 864,714 & 0.2471 & 0.0488 & 0.4923 & 160.93  \\
\hline
Rif-0min & 1 & 960,256 & 0.2823 & 0.0361 & 0.4998 & 10.29 \\
\hline
Rif-0min & 2 & 2,247,295 & 0.2656 & 0.1091 & 0.5047 &  25.08  \\
\hline
Rif-20min & 1 & 1,940,387 & 0.2698 & 0.0820 & 0.5054 & 17.69  \\
\hline
Rif-20min & 2 & 4,229,574 & 0.2153 & 0.1647 & 0.5071 &  14.11 \\
\hline
\end{tabular}  
  \caption{Usual quality control indicators applied to our
    $\sigma^{70}$ samples. PBC stands PCR-bottleneck coefficient, SSD
    for standarized standard deviation, FSR for forward strand ratio,
    NSC for normalized strand cross-correlation coefficient. We
    ommited the RSC (relative strand cross-correlation since ChIP-exo
    experiments don't usually have an acomnying input file)}

  \label{tab:qcbase}
\end{table}

In a good ChIP-Seq experiment the strand cross-correlation is
maximized at the fragment length and there is also a ``phantom peak''
at the sample's read length (\cite{encode_qc} and \cite{strandcc}).
However, in ChIP-exo's case this two peaks are confounded. Hence an
enrichment measure based in the strand cross-correlation such as NSC
or RSC is harder to interpret. In figure \ref{figsup:scc} we plotted
the SCC curves for $\sigma^{70}$ ChIP-exo experiments used to
calculate the NSC and RSC values in table \ref{tab:qcbase}. According
to ENCODE's quality metrics, NSC values close to one corresponds to
low enrichment samples or bad quality datasets, hence in our case the
last four samples would be considered to have higher quality than they
actually have. On the other hand, all PBC values are considerably low
and correspond to samples with severe bottlenecking issues, in
particular we are going to show that for the first two samples we are
obtaining a smaller resolution and identifying a comparable number of
targets as in ChIP-Seq PET experiments.

@@

Figure \ref{fig:qcdiagram} shows a flowchart for the ChIPexoQC
pipeline. Which basically partitions genome by keeping the
non-diggested ChIP-exo regions. Then, for each region it calculates a
series of independent statistics. Finally, it creates several
visualizations designed to asses the overall quality levels of A
ChIP-exo sample.

@@

\color{blue}

\subsubsection{Strand imbalance in ChIP-exo data}
\label{sec:strand_imbalance}

The strand imbalance assesment is based in the observation that the
enriched regions usually have a higher concentration of fragments,
therefore we examine the FSR (defined as the ratio of the number of
forward stranded reads divided by the total number of reads in a given
region) as the region with lower depth are being filtered out. This
indicator is of particular importance, since several methods rely on
the ``peak-pair'' assumption. In table \ref{tab:qcbase}, we calculated
the FSR and noticed that for all the samples, it's value is close to
0.5, which means that there are roughly the same amount of reads in
both strands. To asses

\color{black}

\subsubsection{Enrichment and library complexity in ChIP-exo data}
\label{sec:enri}

@@

We first compared various factors that could affect binding site
identification between ChIP-exo and ChIP-Seq data. in order to compare
distribution of signal and background between ChIP-exo and ChIP-Seq
data, we calculated ChIP tag counts accross the genome by counting the
number of reads mapping to each of \Sexpr{bin_size} non-overlapping
window after extending reads by \Sexpr{frag_len} to their 3' end
directions. ChIP tag counts in ChIP-exo data were linearly related to
ChIP tag counts in ChIP-Seq data for the regions with high ChIP tag
counts (Figure \ref{fig:comp}A). This implies that signals for
pontential binding sites are well reproducible between ChIP-exo and
ChIP-Seq data. On the other hand, there was clear difference in the
background distribution between them. In ChIP-Seq data reads were
almost uniformly distributed over background (non-binding) regions and
the ChIP tag counts in there regions were significantly larger than
zero. In contrast, in ChIP-exo data, there was larger variation in
ChIP tag counts among background regions and ChIP tag counts were much
lower in these regions compared to ChIP-Seq data. There were also
large proportion of regions without any read in ChIP-exo data. These
results indicate that background distribution of ChIP-exo data is less
homogeneous than that of ChIP-Seq data.

\color{blue}

We next evaluated the ``peak pair'' assumption of Rhee and Pugh
\cite{exo1}, MACE \cite{MACE}, CexoR, \cite{cexor} and peakzilla
\cite{peakzilla}, i.e. a peak of reads in the forward strand is
usually paired with a peak of reads in the reverse strand that is
located in the other site of the binding site.

\color{black}

\begin{figure}[h]
  \centering
  \includegraphics[width = .6\textwidth,page = 3 ]{../figs/for_paper/ChIPseqPET_ChIPexo_tagCount_comparison.pdf}
  % \includegraphics[width = .9\textwidth ,page = 3]{../figs/for_paper/MA_plot_fwd_bwd_by_seq.pdf}
  \caption{ A) ChIP signal in ChIP-exo was linearly related to that in
    PET ChIP-Seq data in the region with high ChIP tag counts. In
    contrast, there wer clear differences in their background
    distribution, where several of the background ChIP-exo regions
    were empty.}
  \label{fig:comp}
\end{figure}

@@

\color{red}

@@

\color{black}

@@

\section*{Supplement figures}

\subsection*{Strand cross-correlation for ChIP-exo data}

\begin{figure}[h!]
  \centering
  \includegraphics[width = .8\textwidth]{../figs/for_paper/EColi_strand_cross_corr.pdf}
  \caption{Strand cross-correlation curves for $\sigma^{70}$
    samples.}
  \label{figsup:scc}
\end{figure}

### 2015-11-12 rmarkdown/chip_exo_paper.tex
finished complexity analysis, and added comments to paper

We first compared various factors that could affect binding site
identification between ChIP-exo and ChIP-Seq data. in order to compare
distribution of signal and background between ChIP-exo and ChIP-Seq
data, we calculated ChIP tag counts accross the genome by counting the
number of reads mapping to each of 150 non-overlapping
window after extending reads by 150 to their 3' end
directions. ChIP tag counts in ChIP-exo data were linearly related to
ChIP tag counts in ChIP-Seq data for the regions with high ChIP tag
counts (Figure \ref{fig:comp}A). This implies that signals for
pontential binding sites are well reproducible between ChIP-exo and
ChIP-Seq data. On the other hand, there was clear difference in the
background distribution between them. In CHIP-Seq datam reads were
almost uniformly distributed over background (non-binding) regions and
the ChIP tag counts in there regions were significantly larger than
zero. In contrast, in ChIP-exo data, there was larger variation in
ChIP tag counts among background regions and ChIP tag counts were much
lower in these regions compared to ChIP-Seq data. There were also
large proportion of regions without any read in ChIP-exo data. There
results indicate that background distribution of CHIP-exo data is less
homogeneous than that of ChIP-Seq data.

\begin{figure}[h]
  \centering
  \includegraphics[width = .6\textwidth,page = 3 ]{../figs/for_paper/ChIPseqPET_ChIPexo_tagCount_comparison.pdf}
  \includegraphics[width = .9\textwidth ]{../figs/for_paper/MA_plot_fwd_bwd_by_seq.pdf}
  \caption{Comparison}
  \label{fig:comp}
\end{figure}

>>>>

We first compared various factors that could affect binding site
identification between ChIP-exo and ChIP-Seq data. in order to compare
distribution of signal and background between ChIP-exo and ChIP-Seq
data, we calculated ChIP tag counts accross the genome by counting the
number of reads mapping to each of 150 non-overlapping
window after extending reads by 150 to their 3' end
directions. ChIP tag counts in ChIP-exo data were linearly related to
ChIP tag counts in ChIP-Seq data for the regions with high ChIP tag
counts (Figure \ref{fig:comp}A). This implies that signals for
pontential binding sites are well reproducible between ChIP-exo and
ChIP-Seq data. On the other hand, there was clear difference in the
background distribution between them. In ChIP-Seq data reads were
almost uniformly distributed over background (non-binding) regions and
the ChIP tag counts in there regions were significantly larger than
zero. In contrast, in ChIP-exo data, there was larger variation in
ChIP tag counts among background regions and ChIP tag counts were much
lower in these regions compared to ChIP-Seq data. There were also
large proportion of regions without any read in ChIP-exo data. These
results indicate that background distribution of ChIP-exo data is less
homogeneous than that of ChIP-Seq data.

\color{blue}

We next evaluated the ``peak pair'' assumption of Rhee and Pugh
\cite{exo1}, MACE \cite{MACE}, CexoR, \cite{cexor} and peakzilla
\cite{peakzilla}, i.e. a peak of reads in the forward strand is
usually paired with a peak of reads in the reverse strand that is
located in the other site of the binding site.

\color{black}

\begin{figure}[h]
  \centering
  \includegraphics[width = .6\textwidth,page = 3 ]{../figs/for_paper/ChIPseqPET_ChIPexo_tagCount_comparison.pdf}
  % \includegraphics[width = .9\textwidth ,page = 3]{../figs/for_paper/MA_plot_fwd_bwd_by_seq.pdf}
  \caption{ A) ChIP signal in ChIP-exo was linearly related to that in
    PET ChIP-Seq data in the region with high ChIP tag counts. In
    contrast, there wer clear differences in their background
    distribution, where several of the background ChIP-exo regions
    were empty.}
  \label{fig:comp}
\end{figure}
